---
layout: page
title: Publications
permalink: /publications/
---

Scholary papers describing the methodology
========================================================

```{r setup, include = FALSE, cache = FALSE}
library("RefManageR")
bib <- ReadBib("TH.bib", check = FALSE)
BibOptions(check.entries = FALSE, style = "html", cite.style = "authoryear",
           bib.style = "numeric")
```
The party package implements conditional inference trees
`r AutoCite(bib, "Hothorn:2006:JCGS", before = "function ctree(), ")`,
model-based recursive partitioning 
`r AutoCite(bib, "Zeileis+Hothorn+Hornik:2008", before = "function mob(), ")`,
and conditional inference forests
`r AutoCite(bib, "Hothorn:2006:Biostatistics:16344280", before = "function cforest(), ")`.
Variable importances for conditional inference forests are
implemented in function varimp() as described in
`r Citet(bib, "Strobl:2006:BMC-Bioinformatics")` and
`r Citet(bib, "Strobl_Boulesteix_Kneib_2008")`.
An introduction to trees and forests was published by 
`r Citet(bib, "Strobl_Malley_Tutz_2009")`.

**References**

```{r results = "asis", echo = FALSE}
PrintBibliography(bib, .opts = list(check.entries = FALSE, sorting = "ynt"))
```
